Cluster analysis of butterfly transcriptomes
========================================================

Cluster & analyze only those mapping to an Hme ortholog.

```{r}
library(pheatmap)
library(pvclust)
library(DESeq2)
```
Let's define a function for plotting a heatmap with more descriptive names.

```{r fig.width=14, fig.height=14}
labelled_heatmap <- function(data, meta){
  corrs <- cor(data)
	display.names <- paste(meta$Host_Plant,meta$Tissue,meta$Host.Plant.use,sep=".")
	names(display.names) <- meta$Customer_ID
	colnames(corrs) <- display.names[colnames(corrs)]
	rownames(corrs) <- colnames(corrs)
	pheatmap(corrs, cex=0.8)
}
```
... and some functions for normalization and pre-processing of counts.

```{r}
normalize.voom <- function(counts){
  require(limma)
	return(voom(counts)$E)
}
 
cpm.tmm <- function(counts, groups=NA){
	require(edgeR)
	if(is.na(groups)){
		d<-DGEList(counts=counts)
	}
	else{
		d<-DGEList(counts=counts, group=groups)
	}
	d <- calcNormFactors(d, method="TMM") 
	return(cpm(d, normalized.lib.sizes=TRUE))
}
```
Read count table and metadata (information about samples). We remove the failed sample (B1MT) at once to keep the metadata table "in sync" with the count table - it's more convenient if they inculde exactly the same samples.

```{r}
counts <- read.delim("genewise_readcount2.txt",sep="\t",row.names=1)
meta  <- read.delim("rna-seqecoevowabi_relational_table2.txt")
#meta <- meta[-which(meta$Customer_ID=="B1MT"),]
```
Define some vectors that will be useful for coloring plots later.

```{r}
tissue <- meta$Tissue
names(tissue) <- meta$Customer_ID
hostplant <- meta$Host_Plant
names(hostplant) <- meta$Customer_ID
pgroup <- meta$Phylogeny_group
names(pgroup) <- meta$Customer_ID
type <- meta$Host.Plant.use
names(type) <- meta$Customer_ID
```

Normalize and do a principal component analysis.

```{r:normalize_PCA}
tmm <- cpm.tmm(counts)
labelled_heatmap(tmm,meta)

log.cpm.tmm <- normalize.voom(tmm)
labelled_heatmap(log.cpm.tmm,meta)
p <- prcomp(t(log.cpm.tmm))

```

Visualize two of the components, colored by tissue.

```{r:pca_tissue}
comp1 <- 1
comp2 <- 2
plot(p$x[,c(comp1,comp2)], col=as.numeric(tissue[colnames(counts)]),pch=20,main=paste0("PCs ", comp1, ",", comp2))
legend("topright", legend=c("Fat body","Gut","Labial gland","Malpighian tubules"),pch=20,col=1:4)
```


```{r:pca_four_colorings}
comp1 <- 1
comp2 <- 2
par(mfrow=c(2,2))
# Color by tissue
plot(p$x[,c(comp1,comp2)], col=as.numeric(tissue[colnames(counts)]),pch=20,main=paste0("Tissue, PCs ", comp1, ",", comp2))
legend("topright", legend=c("Fat body","Gut","Labial gland","Malpighian tubules"),pch=20,col=1:4)
# Color by host plant
plot(p$x[,c(comp1,comp2)], col=as.numeric(hostplant[colnames(counts)]),pch=20,main=paste0("Host plant, PCs ", comp1, ",", comp2))
legend("topright", legend=unique(hostplant),pch=20,col=1:length(unique(hostplant)))
# Color by phylogeny group
plot(p$x[,c(comp1,comp2)], col=as.numeric(pgroup[colnames(counts)]),pch=20,main=paste0("Phylogroup, PCs ", comp1, ",", comp2))
legend("topright", legend=unique(pgroup),pch=20,col=1:length(unique(pgroup)))
# Color by core vs extended
plot(p$x[,c(comp1,comp2)], col=as.numeric(type[colnames(counts)]),pch=20,main=paste0("CoreVsExtended, PCs ", comp1, ",", comp2))
legend("topright", legend=unique(type),pch=20,col=1:length(unique(type)))
```

Loop over combinations of PCs and color by family. 
```{r:pca_loop}
par(mfrow=c(4,4))
for (comp1 in 1:4){
  for (comp2 in 1:4){
		if (comp1 != comp2){
	# Color by family
	family = meta$Family
	names(family) <- meta$Customer_ID
	plot(p$x[,c(comp1,comp2)], 	col=as.numeric(family[colnames(counts)]),pch=20,main=paste0("PCs ", comp1, ",", comp2))
#legend("topright", 	legend=unique(family),pch=20,col=1:length(unique(family)))
	}
}
}
```

Let's look at gut samples only, colored by core vs extended.

```{r:pca_gut}
subset <- meta[which(meta$Tissue=="Gut"),"Customer_ID"]
columns <- intersect(subset, colnames(tmm))
x <- tmm[,columns]

x.meta <- meta[which(meta$Tissue=="Labial gland"),]
x.log <- normalize.voom(x)

#p <- prcomp(t(x.log[which(rowMeans(x.log)>1),]))
p <- prcomp(t(x.log))

par(mfrow=c(5,4))
for (comp1 in 1:5){
  for (comp2 in 1:5){
		if (comp1 != comp2){
	plot(p$x[,c(comp1,comp2)], 	col=as.numeric(type[colnames(x)]),pch=20,main=paste0("PCs ", comp1, ",", comp2))
#legend("topright", 	legend=unique(family),pch=20,col=1:length(unique(family)))
	}
}
}
```

PCA of gut samples with more informative text labels, colored by core vs extended.

```{r:pca_gut_textlabels}
par(mfrow=c(1,1))
comp1 <- 1
comp2 <- 2
plot(p$x[,c(comp1,comp2)], 	col=as.numeric(type[colnames(x.log)]),pch=20,main=paste0("PCs ", comp1, ",", comp2))
legend("topright", 	legend=unique(type),pch=20,col=1:length(unique(type)))
display.names <- paste(meta$Host_Plant,meta$Phylogeny_group,meta$Host.Plant.use,sep=".")
names(display.names) <- meta$Customer_ID
display.names.gut <- display.names[colnames(x.log)]
text(p$x[,comp1],p$x[,comp2],labels=display.names.gut,cex=0.6)
```

Try clustering with bootstrapping (pvclust) on gut samples.
We have commented out the actual commands here because they take a long time to run. Instead, we read a previously saved version of the pvclust output.

```{r:pvclust_gut}
# Using contigs with CPM > 1
# res <- pvclust(x.log[which(rowMeans(x)>1),],nboot=100,method.hclust="complete")
# Using all contigs
# res <- pvclust(x.log,nboot=100,method.hclust="complete")
# 1000 bootstrap samples for all contigs, complete linkage
# res <- pvclust(x.log,nboot=1000,method.hclust="complete")
#save(res, file="gut_pvclust_complete_1000.Robj")
#pdf("gut_ortho_pvclust_complete_100.pdf")
#plot(res)
#dev.off()
```

Core vs extended differential gene expression in gut
====================================================

Using only gut samples (n=18).

```{r:deseq}
# DESeq2
subset <- meta[which(meta$Tissue=="Gut"),"Customer_ID"]
columns <- intersect(subset, colnames(counts))
meta.gut <- meta[which(meta$Tissue=="Gut"),]

dds <- DESeqDataSetFromMatrix(countData = counts[,columns], colData = meta.gut[,c("Customer_ID","Host.Plant.use","Phylogeny_group","Host_Plant","Family")], design = ~Family+Host.Plant.use)

dds <- DESeq(dds, betaPrior=FALSE)
res <- results(dds)
sig <- res[which(res$padj<0.01),]
sig.o <- sig[order(sig$padj),]
print(dim(sig.o))
write.table(sig.o,file="sig_gut_corevsextended_DESeq2_0.01.txt",quote=F)
```

Labial gland
-------------
```{r:labi}
subset <- meta[which(meta$Tissue=="Labial gland"),"Customer_ID"]
columns <- intersect(subset, colnames(counts))
meta.gut <- meta[which(meta$Tissue=="Labial gland"),]

dds <- DESeqDataSetFromMatrix(countData = counts[,columns], colData = meta.gut[,c("Customer_ID","Host.Plant.use","Phylogeny_group","Host_Plant","Family")], design = ~Family+Host.Plant.use)

dds <- DESeq(dds, betaPrior=FALSE)
res <- results(dds)
sig <- res[which(res$padj<0.01),]
sig.o <- sig[order(sig$padj),]
print(dim(sig.o))
write.table(sig.o,file="sig_labialgland_corevsextended_DESeq2_0.01.txt",quote=F)
```

Fat body
-------------
```{r:fat}
subset <- meta[which(meta$Tissue=="Fat body"),"Customer_ID"]
columns <- intersect(subset, colnames(counts))
meta.gut <- meta[which(meta$Tissue=="Fat body"),]

dds <- DESeqDataSetFromMatrix(countData = counts[,columns], colData = meta.gut[,c("Customer_ID","Host.Plant.use","Phylogeny_group","Host_Plant","Family")], design = ~Family+Host.Plant.use)

dds <- DESeq(dds, betaPrior=FALSE)
res <- results(dds)
sig <- res[which(res$padj<0.01),]
sig.o <- sig[order(sig$padj),]
print(dim(sig.o))
write.table(sig.o,file="sig_fatbody_corevsextended_DESeq2_0.01.txt",quote=F)
```

Malpighian tubules
-------------
```{r:malpi}
subset <- meta[which(meta$Tissue=="Malpighian tubules"),"Customer_ID"]
columns <- intersect(subset, colnames(counts))
meta.gut <- meta[which(meta$Tissue=="Malpighian tubules"),]

dds <- DESeqDataSetFromMatrix(countData = counts[,columns], colData = meta.gut[,c("Customer_ID","Host.Plant.use","Phylogeny_group","Host_Plant","Family")], design = ~Family+Host.Plant.use)

dds <- DESeq(dds, betaPrior=FALSE)
res <- results(dds)
sig <- res[which(res$padj<0.01),]
sig.o <- sig[order(sig$padj),]
print(dim(sig.o))
write.table(sig.o,file="sig_malpi_corevsextended_DESeq2_0.01.txt",quote=F)
```

DE analysis between core and extended using all samples
==========================================================

... with tissue and family as factors.

```{r:deseq-allsamples}
columns <- intersect(meta$Customer_ID, colnames(counts))
counts_ <- counts[,columns]

dds <- DESeqDataSetFromMatrix(countData = counts_, colData = meta[,c("Customer_ID","Host.Plant.use","Phylogeny_group","Host_Plant","Tissue","Family")], design = ~Tissue+Family+Host.Plant.use)

dds <- DESeq(dds, betaPrior=FALSE, fitType="local")
res <- results(dds)
sig <- res[which(res$padj<0.01),]
sig.o <- sig[order(sig$padj),]
head(sig.o)
print(dim(sig.o))
write.table(sig.o,file="sig_coreextended_alltissues_DESeq2_0.01.txt",quote=F)
```
Plot some examples of genes with low p values. These are differentially expressed in core vs extended overall, that is, considering all tissues.

```{r plot-genes}
library(lattice)

for (gene in head(rownames(sig.o))){
  expr <- t(counts_[gene,]/colSums(counts_))
  temp <- data.frame(expr=expr, tissue=meta$Tissue, core_vs_ext=meta$Host.Plant.use)
  print(bwplot(expr~core_vs_ext|tissue, data=temp, main=gene))
  readline()
}
```

Differential expression between phylogenetic groups
===================================================

Using all samples, with family and tissue as factors.

```{r:plotfcn}
plotphylo <- function(sig, counts){
for (gene in head(rownames(sig))){
  expr <- t(counts[gene,]/colSums(counts))
  temp <- data.frame(expr=expr, tissue=meta$Tissue, phylo=meta$Phylogeny_group)
  print(bwplot(expr~phylo|tissue, data=temp, main=gene))
}
}
```

Rosids vs. Ranunculales
-----------------------

```{r:deseq-phylogroups-1}
columns <- intersect(meta$Customer_ID, colnames(counts))
counts_ <- counts[,columns]

dds <- DESeqDataSetFromMatrix(countData = counts_, colData = meta[,c("Customer_ID","Host.Plant.use","Phylogeny_group","Host_Plant","Tissue","Family")], design = ~Tissue+Family+Phylogeny_group)

dds <- DESeq(dds, betaPrior=FALSE, fitType="local")
#res <- results(dds, contrast=c("Phylogeny_group","Rosids","Asterids"))
res <- results(dds, contrast=c("Phylogeny_group","Rosids","Ranunculales"))
#res <- results(dds, contrast=c("Phylogeny_group","Ranunculales","Asterids"))
sig <- res[which(res$padj<0.01),]
sig.o <- sig[order(sig$padj),]
head(sig.o)
print(dim(sig.o))
plotphylo(sig.o, counts_)
write.table(sig.o,file="sig_rosids_ranunc_alltissues_DESeq2_0.01.txt",quote=F)
```


Rosids vs. Asterids
-----------------------

```{r:deseq-phylogroups-2}
res <- results(dds, contrast=c("Phylogeny_group","Rosids","Asterids"))
sig <- res[which(res$padj<0.01),]
sig.o <- sig[order(sig$padj),]
head(sig.o)
print(dim(sig.o))
plotphylo(sig.o, counts_)
write.table(sig.o,file="sig_rosids_asterids_alltissues_DESeq2_0.01.txt",quote=F)
```

Ranunculales vs. Asterids
-----------------------

```{r:deseq-phylogroups-3}
res <- results(dds, contrast=c("Phylogeny_group","Ranunculales","Asterids"))
sig <- res[which(res$padj<0.01),]
sig.o <- sig[order(sig$padj),]
head(sig.o)
print(dim(sig.o))
plotphylo(sig.o, counts_)
write.table(sig.o,file="sig_ranunc_asterids_alltissues_DESeq2_0.01.txt",quote=F)
```